---
title: "Reproducible Research: Peer Assessment 1"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

## Loading and preprocessing the data

```{r load_data}
# Load the data (activity.csv must be in the working directory)
activity <- read.csv("activity.csv")

# Convert date column from character to Date type
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")

# Inspect the data
head(activity)
str(activity)
summary(activity)
```

The dataset contains **`r nrow(activity)`** observations across three variables:
`steps`, `date`, and `interval`.

---

## What is the mean total number of steps taken per day?

*Missing values (`NA`) are ignored in this section.*

```{r steps_per_day}
library(ggplot2)

# Calculate total steps per day
steps_per_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
```

```{r histogram_steps}
# Histogram of total steps per day
ggplot(steps_per_day, aes(x = steps)) +
  geom_histogram(binwidth = 2000, fill = "steelblue", color = "white") +
  labs(
    title = "Histogram of Total Steps Taken Each Day",
    x     = "Total Steps per Day",
    y     = "Frequency"
  ) +
  theme_minimal()
```

```{r mean_median}
mean_steps   <- mean(steps_per_day$steps)
median_steps <- median(steps_per_day$steps)

cat("Mean total steps per day:  ", round(mean_steps, 2), "\n")
cat("Median total steps per day:", median_steps, "\n")
```

- **Mean** total number of steps taken per day: **`r format(round(mean_steps, 2), big.mark = ",")`**
- **Median** total number of steps taken per day: **`r format(median_steps, big.mark = ",")`**

---

## What is the average daily activity pattern?

```{r time_series}
# Compute average steps per 5-minute interval across all days
interval_avg <- aggregate(steps ~ interval, data = activity, FUN = mean, na.rm = TRUE)

# Time series plot
ggplot(interval_avg, aes(x = interval, y = steps)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  labs(
    title = "Average Number of Steps per 5-Minute Interval",
    x     = "5-Minute Interval",
    y     = "Average Number of Steps"
  ) +
  theme_minimal()
```

```{r max_interval}
max_interval  <- interval_avg$interval[which.max(interval_avg$steps)]
max_steps_val <- round(max(interval_avg$steps), 2)

cat("Interval with maximum average steps:", max_interval, "\n")
cat("Average steps at that interval:     ", max_steps_val, "\n")
```

The 5-minute interval that contains the maximum average number of steps across all days
is interval **`r max_interval`**, with an average of **`r max_steps_val`** steps.

---

## Imputing missing values

```{r count_na}
# Total number of rows with missing step values
total_na <- sum(is.na(activity$steps))
cat("Total number of missing values (rows with NA):", total_na, "\n")
```

The total number of rows with missing values (`NA`) is **`r total_na`**.

**Imputation strategy:** Each missing `steps` value is replaced with the
**mean number of steps for that specific 5-minute interval**, averaged across
all non-missing days. This preserves the typical daily activity rhythm and is
a simple yet effective approach.

```{r impute}
# Build a named lookup: interval -> mean steps
interval_means <- setNames(interval_avg$steps, interval_avg$interval)

# Copy original dataset and fill NAs
activity_imputed <- activity
na_rows <- is.na(activity_imputed$steps)
activity_imputed$steps[na_rows] <- interval_means[
  as.character(activity_imputed$interval[na_rows])
]

# Confirm no NAs remain
cat("Missing values remaining after imputation:", sum(is.na(activity_imputed$steps)), "\n")
```

```{r imputed_histogram}
# Recalculate total steps per day using imputed data
steps_per_day_imp <- aggregate(steps ~ date, data = activity_imputed, FUN = sum)

# Histogram after imputation
ggplot(steps_per_day_imp, aes(x = steps)) +
  geom_histogram(binwidth = 2000, fill = "darkorange", color = "white") +
  labs(
    title = "Total Steps Taken Each Day (After Imputation)",
    x     = "Total Steps per Day",
    y     = "Frequency"
  ) +
  theme_minimal()
```

```{r imputed_mean_median}
mean_imp   <- mean(steps_per_day_imp$steps)
median_imp <- median(steps_per_day_imp$steps)

cat("Mean total steps per day (imputed):  ", round(mean_imp, 2), "\n")
cat("Median total steps per day (imputed):", round(median_imp, 2), "\n")
```

| Metric | Before Imputation | After Imputation |
|--------|:-----------------:|:----------------:|
| Mean   | `r format(round(mean_steps, 2), big.mark = ",")` | `r format(round(mean_imp, 2), big.mark = ",")` |
| Median | `r format(median_steps, big.mark = ",")` | `r format(round(median_imp, 2), big.mark = ",")` |

**Impact of imputation:** Filling missing values with interval means increases
the count of days clustered around the mean — the central bar of the histogram
grows taller. The **mean** is unchanged (filling with means cannot shift the
mean), while the **median** shifts slightly upward to coincide with the mean.
Previously all-`NA` days now carry a realistic daily total, making the
distribution more symmetric.

---

## Are there differences in activity patterns between weekdays and weekends?

*This section uses the imputed dataset.*

```{r weekday_weekend}
# Create a two-level factor: "weekday" or "weekend"
activity_imputed$day_type <- ifelse(
  weekdays(activity_imputed$date) %in% c("Saturday", "Sunday"),
  "weekend",
  "weekday"
)
activity_imputed$day_type <- factor(
  activity_imputed$day_type,
  levels = c("weekday", "weekend")
)

# Average steps per interval and day type
interval_daytype <- aggregate(
  steps ~ interval + day_type,
  data = activity_imputed,
  FUN  = mean
)
```

```{r panel_plot}
# Panel plot: weekday vs weekend
ggplot(interval_daytype, aes(x = interval, y = steps, color = day_type)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ day_type, ncol = 1) +
  scale_color_manual(values = c("weekday" = "steelblue", "weekend" = "darkorange")) +
  labs(
    title = "Average Steps per 5-Minute Interval: Weekdays vs Weekends",
    x     = "5-Minute Interval",
    y     = "Average Number of Steps"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Observations:**

- On **weekdays**, there is a pronounced activity spike around interval 835
  (approximately 8:35 AM), likely corresponding to a morning commute or
  exercise routine, followed by lower, relatively flat activity for the rest
  of the day.
- On **weekends**, activity is more evenly spread across the day, with several
  moderate peaks during daytime hours, reflecting a more flexible and active
  daily pattern.
- Overall, **weekends show higher sustained activity** throughout the day
  compared to weekdays, even though the single highest peak belongs to weekdays.
